---
title: "HDP for mutational signatures analysis"
author: "Nicola Roberts"
date: "April 4, 2016"
output: 
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{HDP for mutational signatures analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()

knitr::opts_chunk$set(fig.width=6, fig.height=5)
```

For the background to this document, see the other package vignette
"General introduction to the hdp package."

One specific application of the HDP is the analysis of somatic mutation data from
cancer genome sequencing projects. In this setting, the input data consists of
counts of mutation categories across a number of cancer samples. The HDP model
returns a set of components representing the underlying mutational processes with
their characteristic distributions or 'signatures' over the set of possible
mutation classes.

The ability to define hierarchies of sample-relatedness (via the tree of parent 
DP nodes in a customised HDP structure) allows mutational signatures to be
inferred both across and within groups of patients, and groups of mutations 
within patients. Patient groups could be defined by cancer type or driver mutation
status, while mutation groups *within* a patient could be defined by temporal or
regional information, subclone status, etc.


# Example: pan-cancer analysis of lung, ovarian and skin cancer

## Tally somatic mutation categories

The `r Biocexptpkg("SomaticCancerAlterations")` package contains several somatic alteration
datasets for different cancer types. In this example, the number of somatic base
substitution mutations in each of the 96 categories defined by local trinucleotide context
are tallied across 100 lung cancers, 100 ovarian cancers, and 100 melanomas. 

```{r loaddata, message=FALSE}
library(hdp)
library(GenomicRanges)

# Lung adenocarcinoma
data(luad_tcga, package="SomaticCancerAlterations")

# Ovarian serous cystadenocarcinoma
data(ov_tcga, package="SomaticCancerAlterations")

# Skin cutaneous melanoma
data(skcm_tcga, package="SomaticCancerAlterations")

# only keep SNP type, add cancer type to sample name, and only keep
# necessary metadata. Only keep 100 samples. Then concatenate and sort.
for (cancer_name in c("luad", "ov", "skcm")){
  raw <- get(paste0(cancer_name, "_tcga"))
  snv <- raw[which(raw$Variant_Type == "SNP")]
  snv <- snv[which(snv$Patient_ID %in% levels(snv$Patient_ID)[1:100])]
  mcols(snv) <- data.frame(sampleID=paste(cancer_name, snv$Patient_ID, sep='_'),
                           ref=snv$Reference_Allele,
                           alt=snv$Tumor_Seq_Allele2)
  assign(cancer_name, snv)
}
variants <- sort(c(luad, ov, skcm))
remove(cancer_name, luad, luad_tcga, ov, ov_tcga, skcm, skcm_tcga, raw, snv)

variants


# tally mutation counts in 96 base substitution classes defined by trinucleotide context
# bit slow - don't run. Load pre-made copy. 
# mut_count <- nrmisc::tally_mutations_96(variants)
dim(mut_count)
head(mut_count[,1:5])
tail(mut_count[,1:5])
```


## Initialise HDP structure

In this example, the HDP is structured to have

* one top grandparent DP drawing from the base distribution (ppindex 0) with 
its own concentration parameter (cpindex 1),
* three parent DPs (one per cancer type) drawing from the grandparent distribution 
(ppindex 1) and sharing a new concentration parameter (cpindex 2),
* 100 child DPs (one per lung cancer sample) drawing from the first parent (ppindex 2)
and sharing a new concentration parameter (cpindex 3),
* 100 child DPs (one per ovarian cancer sample) drawing from the second parent (ppindex 3)
and sharing a new concentration parameter (cpindex 4), and
* 100 child DPs (one per melanoma sample) drawing from the third parent (index 4)
and sharing a new concentration parameter (cpindex 5).

The base distribution is a uniform Dirichlet with pseudocount 1 in each of the 96
possible mutation categories. Each concentration parameter is drawn from a gamma prior
with hyperparameters rate=1, shape=1. 


```{r inithdp}
# initialise HDP
ppindex <- c(0, rep(1, 3), rep(2:4, each=100))
cpindex <- c(1, rep(2, 3), rep(3:5, each=100))
hdp <- hdp_init(ppindex, 
                cpindex, 
                hh=rep(1, 96), 
                alphaa=rep(1, 5), 
                alphab=rep(1, 5))

# add data
hdp <- hdp_setdata(hdp, 5:numdp(hdp), mut_count)

# activate DPs, 10 initial components
hdp <- dp_activate(hdp, 1:numdp(hdp), 10, seed=1)

hdp

```

## Run multiple posterior sampling chains

Run four independent posterior sampling chains with 4000 burn-in iterations,
then collect 50 samples off each chain with 50 iterations between each. 

```{r postsamp}
# Run four independent posterior sampling chains
# Takes ~15 minutes - don't run. Load pre-made copy.
# chlist <- vector("list", 4)
# 
# for (i in 1:4){
#   chlist[[i]] <- hdp_posterior(hdp, 
#                                burnin=4000,
#                                n=50,
#                                space=50,
#                                cpiter=3, 
#                                seed=i*1e4)
# }
# 
# mut_example_multi <- hdp_multi_chain(chlist)
mut_example_multi

par(mfrow=c(2,2), mar=c(4, 4, 2, 1))
p1 <- lapply(chains(mut_example_multi), plot_lik, bty="L", start=500)
p2 <- lapply(chains(mut_example_multi), plot_numcluster, bty="L")
p3 <- lapply(chains(mut_example_multi), plot_data_assigned, bty="L")
```
.

## Extract components (mutational signatures)

The extracted components represent the underlying mutational processes
giving rise to the observed catalogues of somatic mutation. Several processes
are recognisable from [COSMIC](http://cancer.sanger.ac.uk/cosmic/signatures), 
for example (1) is the UV radiation signature dominant in melanomas, (2) is the tobacco
signature dominant in lung cancers, etc.  

The `plot_dp_comp_exposure()` function plots the estimated proportion of mutations 
within each sample derived from each signature. Note that by setting `incl_nonsig=FALSE`, 
only those signatures with non-zero 95% credibility intervals for exposure in a 
sample are included. As a result, a fraction of the sample's signature 
exposure is left unexplained, as we can't have confidence that the other signatures 
truly contribute to that sample. 

```{r extractmulti, fig.width=8, fig.height=5, fig.keep="all"}
mut_example_multi <- hdp_extract_components(mut_example_multi)
mut_example_multi

par(mfrow=c(1,1), mar=c(5, 4, 4, 2))
plot_comp_size(mut_example_multi, bty="L")

bases <- c("A", "C", "G", "T")
trinuc_context <- paste0(rep(rep(bases, times=6), each=4),
                         rep(c("C", "T"), each=48),
                         rep(bases, times=24))
group_factor <- as.factor(rep(c("C>A", "C>G", "C>T", "T>A", "T>C", "T>G"),
                           each=16))
for (i in 0:(length(comp_categ_counts(mut_example_multi)) -1)) {
  print(plot_comp_distn(mut_example_multi, comp=i, cat_names=trinuc_context,
                 grouping=group_factor, col=RColorBrewer::brewer.pal(6, "Set1"),
                 col_nonsig="grey80", show_group_labels=TRUE))
}

plot_dp_comp_exposure(mut_example_multi, main_text="Lung adenocarcinoma",
                      dpindices=4+(1:100), 
                      col=RColorBrewer::brewer.pal(12, "Set3"),
                      incl_nonsig=FALSE)

plot_dp_comp_exposure(mut_example_multi, main_text="Ovarian cancer",
                      dpindices=104+(1:100), 
                      col=RColorBrewer::brewer.pal(12, "Set3"),
                      incl_nonsig=FALSE)

plot_dp_comp_exposure(mut_example_multi, main_text="Melanoma",
                      dpindices=204+(1:100), 
                      col=RColorBrewer::brewer.pal(12, "Set3"),
                      incl_nonsig=FALSE)

plot_dp_comp_exposure(mut_example_multi,
                      dpindices=2:4, incl_numdata_plot=FALSE,
                      col=RColorBrewer::brewer.pal(12, "Set3"),
                      incl_nonsig=FALSE,
                      dpnames=c("Lung Adeno", "Ovarian", "Melanoma"))
```


# Example: lung squamous cell carcinoma, conditioned on previous dataset 

## Tally somatic mutation categories

Load dataset of mutation counts from 100 lung squamous cell carcinomas. 

```{r loadlusc}
# # bit slow - don't run. Load pre-made copy. 

# data("lusc_tcga", package="SomaticCancerAlterations")
# 
# lusc <- lusc_tcga[which(lusc_tcga$Variant_Type == "SNP")]
# lusc <- lusc[which(lusc$Patient_ID %in% levels(lusc$Patient_ID)[1:101])]
# mcols(lusc) <- data.frame(sampleID=paste('lusc', lusc$Patient_ID, sep='_'),
#                           ref=lusc$Reference_Allele,
#                           alt=lusc$Tumor_Seq_Allele2)
# 
# remove(lusc_tcga)

# # tally mutation counts in 96 base substitution classes defined by trinucleotide context

# lusc_count <- nrmisc::tally_mutations_96(lusc)
# # remove sample with the largest burden (very different to others)
# lusc_count <- lusc_count[-which.max(rowSums(lusc_count)),]
dim(lusc_count)
head(lusc_count[,1:5])
```


## Extend previous HDP structure, freeze old nodes, run posterior sampling chains

Condition on the final `hdpState` from the end of each previous chain by adding 
new nodes for the lung squamous data, freezing the old nodes (except the top 
grandparent DP node), and running additional posterior sampling chains over the 
new nodes. 

```{r addlusc}
# # bit slow - don't run. Load pre-made copy. 
# 
# hdpStatelist <- lapply(chains(mut_example_multi), final_hdpState)
# chlist <- vector("list", 4)
# 
# for (i in 1:4){
#   hdp <- hdpStatelist[[i]]
#   hdp <- hdp_addconparam(hdp, 1, 1)
#   hdp <- hdp_adddp(hdp, 101, 
#                    ppindex=c(1, rep(305, 100)),
#                    cpindex=c(2, rep(6, 100)))
#   hdp <- hdp_setdata(hdp, 306:405, lusc_count)
#   hdp <- dp_freeze(hdp, 2:304)
#   hdp <- dp_activate(hdp, 305:405, initcc=base(hdp)@numclass, seed=i*1e5)
# 
#   chlist[[i]] <- hdp_posterior(hdp, burnin=1500, n=50, 
#                                space=50, cpiter=3, seed=i*1e6)
# }
# 
# lusc_multi <- hdp_multi_chain(chlist)
lusc_multi

par(mfrow=c(2,2), mar=c(4, 4, 2, 1))
p1 <- lapply(chains(lusc_multi), plot_lik, bty="L")
p2 <- lapply(chains(lusc_multi), plot_numcluster, bty="L")
p3 <- lapply(chains(lusc_multi), plot_data_assigned, bty="L")
```


## Extract components

```{r lusccomp, fig.width=8, fig.height=5}
lusc_multi <- hdp_extract_components(lusc_multi)
lusc_multi

plot_comp_size(lusc_multi, bty="L")

for (i in 0:(length(comp_categ_counts(lusc_multi)) -1)) {
  print(plot_comp_distn(lusc_multi, comp=i, cat_names=trinuc_context,
                 grouping=group_factor, col=RColorBrewer::brewer.pal(6, "Set1"),
                 col_nonsig="grey80", show_group_labels=TRUE))
}

plot_dp_comp_exposure(lusc_multi, main_text="Lung squamous",
                      dpindices=306:405, 
                      col=RColorBrewer::brewer.pal(12, "Set3"),
                      incl_nonsig=FALSE)

plot_dp_comp_exposure(lusc_multi,
                      dpindices=c(305, 2:4), incl_numdata_plot=FALSE,
                      col=RColorBrewer::brewer.pal(12, "Set3"),
                      incl_nonsig=FALSE,
                      dpnames=c("Lung Squamous", "Lung Adeno", 
                                "Ovarian", "Melanoma"))

```

. 


# Example: lung squamous cell carcinoma, conditioned on signatures database

Use `hdp_prior_init` to condition on previously identified signatures 
(e.g. COSMIC database). 


```{r prior}
cosmic.sigs <- read.table('http://cancer.sanger.ac.uk/cancergenome/assets/signatures_probabilities.txt', header=TRUE, sep='\t')
#  sort by Substitution Type and Trinucleotide
cosmic.sigs <- cosmic.sigs[order(cosmic.sigs$Substitution.Type, cosmic.sigs$Trinucleotide),]
sigs <- as.matrix(cosmic.sigs[,grep('Signature', colnames(cosmic.sigs))])
prior_pseudoc <- rep(100, 30)
lusc_prior <- hdp_prior_init(sigs, prior_pseudoc, hh=rep(1, 96),
                             alphaa=c(1, 1), alphab=c(1, 2))

lusc_prior <- hdp_addconparam(lusc_prior, c(1,1), c(1,1))
lusc_prior <- hdp_adddp(lusc_prior, 101,
                  ppindex=c(1, rep(1+30+1, 100)),
                  cpindex=c(3, rep(4, 100)))
lusc_prior <- hdp_setdata(lusc_prior, (1+30+1)+1:100, lusc_count)

chlist <- vector("list", 4)
for (i in 1:4){
      lusc_pr <- dp_activate(lusc_prior, 
                                       (1+30+1)+0:100, initcc=30)
      chlist[[i]] <- hdp_posterior(lusc_pr, burnin=1500, n=50, 
                               space=50, cpiter=3, seed=i*1e6)
}

lusc_pr_multi <- hdp_multi_chain(chlist)
lusc_pr_ec <- hdp_extract_components(lusc_pr_multi)

plot_dp_comp_exposure(lusc_pr_ec, 1+30+1+(1:100), incl_nonsig = FALSE,
                      col=rep(nrmisc::nr_col_pal("bigpal", 29), 2))
```


# Session info

Session information for the system on which this document was compiled:

```{r sessionInfo, echo=TRUE}
devtools::session_info()
```
